---
title: "Week 2: R Exercise and Homework"
output: github_document
---

# Analysis Routine

On August 29, DJ Patil, the Chief Data Scientist in the White House Office
of Science and Technology Policy, and Mark Rosekind, the Administrator of
the National Highway Traffic Safety Administration (NHTSA) announced the
release of a data set documenting all traffic fatalities occurring in the
United States in 2015. In this excise, we tried to use the command we have
learned to analyze this dataset.

## Analysis Routine: Data Import

- To get started with this exercise, direct your R session to a dedicated
workshop directory and download into this directory the following sample
tables. Then import the files into Excel and save them as tab delimited
text files.

    - [accident](./accident.csv)
    - [Description of the fields in the dataset](ftp://ftp.nhtsa.dot.gov/FARS/FARS-DOC/Analytical User Guide/USERGUIDE-2015.pdf)

- Import the tables into R

```{r import_data1, eval=T}
accidents <- read.delim(file="accident.csv", header=T, sep=",") 
head(accidents)
names(accidents)
``` 
- Preprocess the data

```{r import_data2, eval=T}
#Only include data with correct long/lat coordinates and valid hours
accidents <- subset(accidents, LONGITUD!=999.99990 &  LONGITUD!=888.88880 & LONGITUD!=777.77770 & HOUR<30)
``` 

- Aggregate the number of accidents by month 

```{r import_data3, eval=T}
hist(accidents$MONTH)
``` 

- Aggregate the number of accidents by hour 

```{r import_data4, eval=T}
hist(accidents$HOUR)
``` 

- Aggregate the number of persons by month

```{r import_data5, eval=T}
a<-aggregate(PERSONS~MONTH, accidents,sum)
plot(a$MONTH, a$PERSONS, type='o', col='blue')
``` 

- Aggregate the number of persons by hour

```{r import_data6, eval=T}
a<-aggregate(PERSONS~HOUR, accidents,sum)
plot(a$HOUR,a$PERSONS, type='o', col='blue' )
``` 

# Homework

Based on the `accident` dataset, please write a function (see the
							  [RBascis.md](./RBasics.md)),
say named `get-state-fatality(stateName, display)`, to return the monthly
fatalites of at state `stateName` and if `display=TRUE` also to plot the
fatalities by months. 


