---
title: "Week 4: Geostatistics I"
output: github_document
---

```{r global_options, echo=FALSE, results='asis'}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',  warning=FALSE, message=FALSE)
rm(list=ls())
library(sp)
library(gstat)
library(RColorBrewer)
library(plotGoogleMaps)
```
# Data

The `meuse` data set contains the measurements of the concentration of
heavy metals around the banks of the Maas river in the area west of the
town Stein in the Netherlands. These heavy metals were accumulated over the
years because of river pollution. The data sets has many variables but here
we will use the coordinates and the concentration of zinc at each data
point. You can see the maps of the study area. 

```{r meuse, echo=T, eval=T}
data(meuse)
coordinates(meuse)=~x+y
proj4string(meuse) = CRS('+init=epsg:28992')
plotGoogleMaps(meuse)
```

We first explored the data a little more to get more comfortable with the
datasets. 

```{r meuse2, echo=T, eval=T}

#check the name of each field
names(meuse)

hist(meuse$zinc,breaks=20)
#shapiro.test(meuse$zinc)

#hist(log(meuse$zinc),breaks=20)
#shapiro.test(log(meuse$zinc))

bubble(meuse, 'zinc', col=brewer.pal(2, 'Reds'), main= 'zinc concentrations (ppm)')

plot(log(zinc)~sqrt(dist), meuse)
abline(lm(log(zinc)~sqrt(dist), meuse))
```
We can first use IDW (inverse distance weighted) method to interpolate the
scattering measurements into a continuous surface. 

```{r idw, echo=T, eval=T}
zinc.idw = idw(zinc~1, meuse, meuse.grid)
#image(meuse.grid['dist'], col=rev(brewer.pal(9, 'Reds')))

str(zinc.idw)
plot(zinc.idw[,1], main='zinc, inverse distance weighted estimation')
```

# Kriging

In addition to the point measurements of in `meuse`, we have another set of
auxiliary dataset `meuse.grid`, containing a set of gridded auxiliary
information about this area of study. We will use the distance to the
river. 

```{r meusegrid, echo=T, eval=T}
data(meuse.grid)
names(meuse)

coordinates(meuse.grid)=~x+y
gridded(meuse.grid)=TRUE 
plot(meuse.grid['dist'], col=rev(brewer.pal(9, 'Reds')), main='distance to river (m)')

```

##Variogram

```{r variogram, echo=T, eval=T}
lzn.vgm = variogram(log(zinc)~1, meuse)
lzn.vgm
plot(lzn.vgm)


lzn.vgm.map = variogram(log(zinc)~1, meuse, map=TRUE, cutoff=1500, width=100)
plot(lzn.vgm.map)


lzn.fit = fit.variogram(lzn.vgm, model = vgm(1, "Sph", 900, 1))
lzn.fit
plot(lzn.vgm, lzn.fit)


# accounting for other covariates/auxiliary information
# residual variogram of the linear regression
lznr.vgm = variogram(log(zinc)~sqrt(dist), meuse)
lznr.fit = fit.variogram(lznr.vgm, model = vgm(1, "Exp", 300, 1))
lznr.fit
plot(lznr.vgm, lznr.fit)
```

### Directional variogram

```{r variogram2, echo=T, eval=T}
lzn.dir = variogram(log(zinc)~1, meuse, alpha = c(0, 45, 90, 135))
lzndir.fit = vgm(.59, "Sph", 1200, .05, anis = c(45, .4))
plot(lzn.dir, lzndir.fit, as.table = TRUE)

lznr.dir = variogram(log(zinc)~sqrt(dist), meuse, alpha = c(0, 45, 90, 135))
plot(lznr.dir, lznr.fit, as.table = TRUE)
```


## Kriging prediction

```{r kriging, echo=T, eval=T}
lzn.kriged = krige(log(zinc)~1, meuse, meuse.grid, model = lzn.fit)
names(lzn.kriged)
plot(lzn.kriged[,1], main='zinc, simple kriging estimation')

#kriging also output the variance of the estimation
plot(lzn.kriged[,2], main='zinc, variance of simple kriging estimation')

# accounting for other covariates/auxiliary information

lznr.kriged = krige(log(zinc)~sqrt(dist), meuse, meuse.grid, model = lzn.fit)

plot(lznr.kriged[,1], main='zinc, universal kriging estimation')

#kriging also output the variance of the estimation
plot(lznr.kriged[,2], main='zinc, variance of universal kriging estimation')
```


### Cross-validation 

```{r kriging, echo=T, eval=T}
choose100 <- sample(1:155, 100) 
training <- meuse[choose100, ] 
validation <- meuse[-choose100, ] 

#Now, we can use the part model to estimate the variogram.
lzn.vgm.training = variogram(log(zinc)~1, training)
lzn.fit.training = fit.variogram(lzn.vgm.training, model = vgm(1, "Sph", 900, 1))
lzn.kriged.training = krige(log(zinc)~1, data=training, newdata=validation, model = lzn.fit.training)
difference <- log(validation$zinc) - lzn.kriged.training$var1.pred 
summary(difference)
rmse <- sqrt(sum(difference^2))
rmse
plot(log(validation$zinc), lzn.kriged.training$var1.pred, xlab= 'observed values', ylab='predicted values')
#abline(lm(log(validation$zinc)~lzn.kriged.training$var1.pred))


#A more automated way is to use the function krige.cv: 
cv<-krige.cv(log(zinc)~1,meuse, model=lzn.fit,nfold=nrow(meuse)) 
names(cv)
rmse2 <- sqrt(sum(cv$residual^2))
rmse2
```

##Simulation: Explore the space of uncertainty

### Unconditional simulation

Generate a set of unconditional simulation based on the specified variogram

```{r simulation1, echo=T, eval=T}
# Generate a set of unconditional simulation based on the specified variogram
xy <- expand.grid(1:100, 1:100)
names(xy) <- c('x','y')
g.dummy <- gstat(formula=z~1, locations=~x+y, dummy=T, beta=1, model=vgm(psill=0.025, range=5, model='Exp'), nmax=20) 
yy <- predict(g.dummy, newdata=xy, nsim=4)
coordinates(yy)=~x+y
gridded(yy) =TRUE
spplot(yy)
```

### Conditional simulation

Generate a set of simulation based on the specified variogram and
conditional to a set of observations

```{r simulation2, echo=T, eval=T}
lzn.condsim = krige(log(zinc)~1, meuse, newdata=meuse.grid, model = lzn.fit, nmax = 30, nsim = 4)
spplot(lzn.condsim, main='four realizations of conditional simulation')

#accounting for covariates
lzn.condsim2 = krige(log(zinc)~sqrt(dist), meuse, newdata=meuse.grid, model = lznr.fit, nmax = 30, nsim = 4)
spplot(lzn.condsim2, main='four realizations of conditional simulation')
```

