--
title: "Week 3: Use R as GIS"
output: github_document
---

```{r global_options, results='asis', warning=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',  warning=FALSE, message=FALSE)
```

```{r load, echo=T, eval=T}
#rm(list=ls())
x <- c("sp", "rgdal", "rgeos", "maptools", "classInt", "RColorBrewer", "GISTools", "maps", "raster", 'ggmap')
install.packages(x) # warning: this may take a number of minutes
lapply(x, library, character.only = TRUE) #load the required packages
```

# Spatial Objects

|       |  Without attributes | With attributes |
| ----- | ------------------  | --------------  |
|Points |  SpatialPoints      | SpatialPointsDataFrame|
|Lines |  SpatialLines      | SpatialLinesDataFrame|
|Polygons |  SpatialPolygons      | SpatialPolygonsDataFrame|
|Raster |  SpatialGrid      | SpatialGridDataFrame|
|Raster |  SpatialPixels      | SpatialPixelsDataFrame|

```{r load_library, echo=T, eval=T}
LubbockBlock<-readShapePoly("Data/LubbockBlockNew.shp") #read polygon shapefile
class(LubbockBlock)
HouseLocation<-read.csv("Data/HouseLatLon.csv") #read GPS data
class(HouseLocation)
coordinates(HouseLocation)<-c('Lon', 'Lat')
class(HouseLocation)
cropland<-raster("Data/Lubbock_CDL_2013_USDA.tif")
class(cropland)

tmin <- getData("worldclim", var = "tmin", res = 10)  # this will download 
class(tmin)
```


```{r load_library1, echo=T, eval=T}
LubbockBlock<-readOGR("./Data", "LubbockBlockNew") #read polygon shapefile
class(LubbockBlock)
```


# Mapping with R

## Basic Mapping

```{r mapping, echo=T, eval=T}
LubbockBlock<-readShapePoly("Data/LubbockBlockNew.shp") #read polygon shapefile
plot(LubbockBlock,axes=TRUE, col=alpha("gray70", 0.6)) #plot Lubbock block shapefile
#add title, scalebar, north arrow, and legend
HouseLocation<-read.csv("Data/HouseLatLon.csv") #read GPS data
price<-HouseLocation$TotalPrice
nclr<-5
priceclr<-brewer.pal(nclr, "Reds")
class<-classIntervals(price, nclr, style="quantile")
clocode<-findColours(class, priceclr)

points(HouseLocation$Lon, HouseLocation$Lat, pch=19, col=clocode, cex=0.5) #add houses on top of Lubbock block shapefile
title(main="Houses on Sale in Lubbock, 2014") 

legend(-101.95, 33.65, legend=names(attr(clocode, "table")), fill =attr(clocode, "palette"), cex=0.5, bty="n")
map.scale(x=-101.85, y=33.49,0.001,"Miles",4,0.5,sfcol='red')
north.arrow(xb=-101.95, yb=33.65, len=0.005, lab="N")

#plot raster
plot(cropland)
#plot raster stack
tmin <- getData("worldclim", var = "tmin", res = 10)  # this will download 
plot(tmin)
```

## Mapping with static Google Maps

```{R mapping2, echo=T, eval=T}
library(RgoogleMaps)
lubbock=geocode('lubbock')

newmap <- GetMap(center = c(lubbock$lat, lubbock$lon), zoom = 12, destfile = "newmap.png", maptype = "roadmap") 

PlotOnStaticMap(newmap, lat=HouseLocation$Lat, lon=HouseLocation$Lon, col='red')
lubbock<-SpatialPolygons(LubbockBlock@polygons, proj4string=CRS("+init=EPSG:4326"))
PlotPolysOnStaticMap(newmap, lubbock, col=alpha('blue', 0.2))
```

## Mapping with dynamic Google Maps

```{R mapping3, echo=T, eval=F}
library(plotGoogleMaps)

data(meuse)
coordinates(meuse)=~x+y
proj4string(meuse) = CRS('+init=epsg:28992')
plotGoogleMaps(meuse, filename='meuse.html')

HouseLocation<-read.csv("Data/HouseLatLon.csv") #read GPS data
coordinates(HouseLocation)<-c('Lon', 'Lat')
proj4string(HouseLocation)=CRS('+init=EPSG:4326')
plotGoogleMaps(HouseLocation, filename='house.html')

ic = iconlabels(meuse$zinc, height=12)
plotGoogleMaps(meuse, iconMarker=ic, mapTypeId='ROADMAP', filename='meuse2.html')

#plot raster
data(meuse.grid)
coordinates(meuse.grid)<-c('x', 'y')
meuse.grid<-as(meuse.grid, 'SpatialPixelsDataFrame')
proj4string(meuse.grid) <- CRS('+init=epsg:28992')
mapMeuseCl<- plotGoogleMaps(meuse.grid,zcol= 'dist',at=seq(0,0.9,0.1),colPalette= brewer.pal(9,"Reds"), filename='meuse3.html') 

#plot polygons
proj4string(LubbockBlock)=CRS("+init=epsg:4326")
m<-plotGoogleMaps(LubbockBlock,zcol="Pop2010",filename= 'MyMap6.htm' , mapTypeId= ' TERRAIN ' ,colPalette= brewer.pal(7,"Reds"), strokeColor="white")

#plot line
meuse.grid<-as(meuse.grid, 'SpatialPixelsDataFrame')
im<-as.image.SpatialGridDataFrame(meuse.grid[ 'dist' ]) 
cl<-ContourLines2SLDF(contourLines(im))
proj4string(cl) <- CRS( '+init=epsg:28992')
mapMeuseCl<- plotGoogleMaps(cl,zcol= 'level' ,strokeWeight=1:9, filename= 'myMap5.htm',mapTypeId= 'ROADMAP') 
```

## Changing map projections

```{r projection, eval=T }

#project a vector 

boudary=readShapePoly('Data/boundary');
proj4string(boudary) <-CRS("+proj=utm +zone=17 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
proj4string(boudary)
boudaryProj<-spTransform(boudary, CRS("+init=epsg:3857"))
proj4string(boudaryProj)


#project a raster
proj4string(cropland)
plot(cropland)
aea <- CRS("+init=ESRI:102003")   #Albert equal area
projCropland=projectRaster(cropland, crs=aea)
plot(projCropland)
```

# Spatial analysis with R


## vector analysis (overlay)

```{r vector, echo=T, eval=T }
#project a vector 

# Datasets
#  * CSV table of (fictionalized) brown bear sightings in Alaska, each
#    containing an arbitrary ID and spatial location specified as a
#    lat-lon coordinate pair. 
#  * Polygon shapefile containing the boundaries of US National Parks
#    greater than 100,000 acres in size. 

bears <- read.csv("Data/bear-sightings.csv")
coordinates(bears) <- c("longitude", "latitude")

# read in National Parks polygons
parks <- readOGR("Data", "10m_us_parks_area")

# tell R that bear coordinates are in the same lat/lon reference system as the parks data 
proj4string(bears) <- proj4string(parks)

# combine is.na() with over() to do the containment test; note that we
# need to "demote" parks to a SpatialPolygons object first
inside.park <- !is.na(over(bears, as(parks, "SpatialPolygons")))

# calculate what fraction of sightings were inside a park
mean(inside.park)
## [1] 0.1720648

# determine which park contains each sighting and store the park name as an attribute of the bears data
bears$park <- over(bears, parks)$Unit_Name

# draw a map big enough to encompass all points, then add in park boundaries superimposed upon a map of the United States
plot(bears)
map("world", region="usa", add=TRUE)
plot(parks, border="green", add=TRUE)
legend("topright", cex=0.85, c("Bear in park", "Bear not in park", "Park boundary"), pch=c(16, 1, NA), lty=c(NA, NA, 1), col=c("red", "grey", "green"), bty="n")
title(expression(paste(italic("Ursus arctos"), " sightings with respect to national parks"))) 

# plot bear points with separate colors inside and outside of parks
points(bears[!inside.park, ], pch=1, col="gray")
points(bears[inside.park, ], pch=16, col="red")

# write the augmented bears dataset to CSV
write.csv(bears, "bears-by-park.csv", row.names=FALSE)

# ...or create a shapefile from the points
writeOGR(bears, ".", "bears-by-park", driver="ESRI Shapefile")
```

## Raster analysis

```{r raster, eval=T, echo=T}
elevation <- getData("alt", country = "ESP")
slope <- terrain(elevation, opt = "slope")
aspect <- terrain(elevation, opt = "aspect")
hill <- hillShade(slope, aspect, 40, 270)
plot(hill, col = grey(0:100/100), legend = FALSE, main = "Spain")
plot(elevation, col = rainbow(25, alpha = 0.35), add = TRUE)

#contours

contour(elevation)
```

```{r raster2, eval=F, echo=T}
#crop raster
plot(hill, col = grey(0:100/100), legend = FALSE, main = "Spain")
plot(elevation, col = rainbow(25, alpha = 0.35), add = TRUE)
extent=drawExtent()
cropElev <- crop(elevation, extent)
plot(cropElev)
```
