--
title: "Homework 5: Writing a Kriging function"
output: github_document
---
```{r global_options, echo=FALSE, results='asis'}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',  warning=FALSE, message=FALSE)
rm(list=ls())
```

The following simple table contains measurements at $10$ locations $z(s_i),
i = 1,\dots 10$, and we need to make estimation of a target location
($z(s_11)$, red point) using kriging.

```{r hw, eval=T, echo=F}

#oa<-readShapePoints('Data/oa.shp')
#oa<-subset(oa, oa$WTE2013>0)
#inc <- sample(1:nrow(oa), 10)
#x <- oa@coords[inc,1]
#y <- oa@coords[inc,2]
#value <- oa$WTE2013[inc]
#obs<-cbind(x,y,value)
#rownames(obs)=1:10
#write.csv(obs, 'data.csv', row.names=F)

obs=read.csv('data.csv')
xtarget=4540000
ytarget=20640000
kable(obs)
plot(obs$x,obs$y)
points(xtarget,ytarget, col='red', pch='*', cex=3)
```

Assume that we know the mean of the observations is $0$. As mentioned in
the class, simple kriging can be used to make this estimation. We have
learned that, with a specification of covariogram/variogram $\gamma(h)$,
the kriging weights can be obtained by $\omega=\sum_{dd}^{-1}\sigma_{dt}$,
where the $\omega$ is the vector of kriging weights, $\sum_{dd}$ is the
covariance matrix of the observation specified by $\gamma(h)$, and
$\sigma_{dt}$ indicates the covariance vector between the targeted location
and the observations. The kriging estimation can then been given as
$\hat{z}(s_{11})=\sum_{i=1}^{10}\omega_{i}z(s_i)$ and the associated
variance is $\hat{\sigma}(s_{11})=\sigma(0) -
\sum_{i=1}^{10}\omega_{i}\sigma(|s_{11}-s_{i}|)$. With these, please finish
the following two questions. 

# Question 1

The exponential covariogram function $\gamma(h)$ is given in the following
R block.

```{r hw2, eval=T, echo=T}

# exponential covariogram
expVariogram<-function(distance){
    nugget=0; sill=2000; range=20000
    return (nugget+sill*exp(-3*distance/range))
}
```

With this covariogram, please finish the following `kriging` function. The
input of this function are observed locations and values (`object'), and
the location ('location')to be estimated. The expected return value of this
function is kriging estimation at `location`.

```{r hw3, eval=T, echo=T}

#With the above exponential covariogram, please finish the following kriging function

kriging<-function(object, location){
	n=nrow(object)
	x <- as.matrix(cbind(object$x, object$y))
	# First, we calculate the distance matrix between observations. 
	# Initialize distance matrix
	x1 <- rep(rep(0,n),n)         
	dist <- matrix(x1,nrow=n,ncol=n) 
	for (i in 1:n){ 
		for (j in 1:n){ 
			dist[i,j]=sqrt((x[i,1]-x[j,1])^2+(x[i,2]-x[j,2])^2) 
		} 
	}
	#######################################################
	## Here, please add your code to convert the distance matrix to covariance
	## matrix


        ## 
	######################################################


	dist2= rep(0,n) # initialize observation to target distance 
	for(i in 1:n){
		dist2[i]=sqrt((x[i,1]-location[1])^2+(x[i,2]-location[2])^2)
	}

	#######################################################
	## Here, please add your code to convert the distance vector between target
	# and observations to the vector of covariance 

        #
	#######################################################

	# The following code generate the kriging weighs for each observations. 
	# function `solve(C)` to invert a matrix `C`, and function `%*%` to
	# multiply two matrix

	w <- solve(C) %*% c

	#######################################################
	# Here, according to the equation given previously in this document,
	# please add your code to get the kriging estimation (say it is called zhat) and the associated
	# variance based on the weight `w`

	zhat <- 0 # Initilized as 0

        #
	#######################################################

	return (zhat)
}
```

# Question 2

Using the `kriging` function, you completed in previous question, please
make estimation at the $10$ observed locations and compare the difference
between your estimation and the observations. What do you find from the
difference?



